---
title: "Data Cleaning"
author: "Padraic O'Hora"
date: "19/11/2024"
output: pdf_document
---
```{r}
if (!requireNamespace("gridExtra", quietly = TRUE)) {
  install.packages("gridExtra")
}
if (!requireNamespace("treemap", quietly = TRUE)) {
  install.packages("treemap")
}

if (!requireNamespace("waffle", quietly = TRUE)) {
  install.packages("waffle")
}

if (!requireNamespace("caret", quietly = TRUE)) {
  install.packages("caret")
}
```

# Setup
```{r include=FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)
library(gridExtra)
library(treemap)
library(waffle)
library(forecast)
library(caret)
```


# data exploration
```{r include=FALSE}

data <- read.csv("./Daily_Household_Transactions.csv")

head(data)
str(data)
summary(data)
```

```{r}
str(data)
```


```{r}
summary(data)
```

It appears some data types are not correct. I will now change the types
```{r}
length(unique(data$Mode)) # 12
length(unique(data$Category)) #50
length(unique(data$Subcategory)) #91
length(unique(data$Income.Expense)) #3 
length(unique(data$Currency)) #1

data$Mode <- as.factor(data$Mode)
data$Category <- as.factor(data$Category)
data$Subcategory <- as.factor(data$Subcategory)
data$Income.Expense <- as.factor(data$Income.Expense)

data <- data %>%
  mutate(Subcategory = ifelse(Subcategory == "", paste(Category, "unknown"), paste(Subcategory)))

data$Date <- parse_date_time(data$Date, orders = c("d/m/Y H:M:S", "d/m/Y"))


```

```{r}
summary(is.na(data))
```

No na values in the data set

# Feature Engineering

## Time features

### Month and Year combinations
```{r}
# Create a new column for month-year
data$MonthYear <- format(data$Date, "%Y-%m")
data$Month <- format(data$Date, "%m")
data$Year <- format(data$Date, "%Y")
head(data)
```

### seasons and year 
```{r}
# Create a new column for month-year
data$Season <- quarter(data$Date , with_year = FALSE)
data$SeasonYear <- quarter(data$Date , with_year = TRUE)
```

### week and year
```{r}
# Create weekly variable
data$Week <- week(data$Date)

# Combine year and week 
data$WeekYear <- paste0(format(data$Date, "%Y"), "-", sprintf("%02d", data$Week))

# Find the minimum year in the dataset
min_year <- min(data$Year)

# Create a new column 'year_number' that maps each year to a sequential number starting from 1
data$year_number <- (as.numeric(data$Year) - as.numeric(min_year)) + 1

# Create a new weeknumber variable that combines year and week
data$weeknumber <- data$year_number * data$Week

# If you want weeknumber as a numeric value, you can convert it
data$weeknumber <- as.numeric(data$weeknumber)

```

Next I will perform univariate analysis

### Exploration of Amount
``` {r}
summary(data$Amount)
```
Median Amount is 100, whereas mean is 2751, suggesting a heavily squewed distribution. 
Max amount is 250000 and min amount is 2
q1 is 35, q3 is 799
IQR is 764

# Univariate Charts

## Amount

### Amount Histogram and boxplot
```{r}
hist(data$Amount)
```
``` {r}
boxplot(data$Amount)
```
#### Chart analysis
both charts show that the amount variable has a considerable amount of outlyer values and is has a long tail over higher values

### Log of Amount Histogram and boxplot
```{r}
data$Amount_log <- log(data$Amount)
hist(data$Amount_log)
```

```{r}
boxplot(data$Amount_log)
```
#### Chart analysis
Using logorythmic scale of the amout variable give a normal distribution 

## Income.Expense 

### Income.Expense Bar chart
```{r}
income_expense_count <- data %>%
  group_by(Income.Expense) %>%
  summarise(Count = n())

ggplot(income_expense_count, aes(x = Income.Expense, y = Count, fill = Income.Expense)) +
  geom_col() +
  geom_text(aes(label = Count), 
            position = position_stack(vjust = 0.5), 
            color = "black", 
            size = 6) +
  labs(title = "Number of Transactions per Income.Expense Category",
       x = "Income.Expense",
       y = "Number of Transactions") 

```
```{r}
stacked_bar_plot <- data %>%
  group_by(Income.Expense) %>%
  summarise(Amount = sum(Amount)) %>%
  ggplot(aes(x = Income.Expense, y = Amount, fill = Income.Expense)) +
  geom_col() +
  labs(title = "Bar Plot of total Amount per Income/Expense ",
       x = "Income/Expense",
       y = "Number of Transactions") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels to 45 degrees
        plot.title = element_text(hjust = 0.5))

stacked_bar_plot


```
#### Chart analysis
Chart shows thhe three subtypes of Income.Expense, Income, Expense and Transfer-out
The dataset consists mostly of Expense observations, with 2000+ entries compared to 
100+ for both income and Transer-out

## Category
### Category Area Chart
```{r}
category_expense_count <- data %>%
  group_by(Category) %>%
  summarise(Count = n())

treemap(category_expense_count,
          index="Category",
          vSize="Count",
          type="index"
          )

```
#### Chart analysis
Chart shows there are many categories in the dataset. The most common category is 
Food, followed by Transportation, Household and Subscription.

## Mode
### Mode Bar Chart
```{r}
mode_expense_count <- data %>%
  group_by(Mode) %>%
  summarise(Count = n())

ggplot(mode_expense_count, aes(x = Mode, y = Count, fill = Mode)) +
  geom_col() +
  geom_text(aes(label = Count), 
            position = position_stack(vjust = 0.5), 
            color = "black", 
            size = 6) +
  labs(title = "Number of Transactions per Mode Category",
       x = "Mode",
       y = "Number of Transactions") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```
#### Chart analysis
Chart shows there are two main Modes of transaction. Savings bank account 1, and Savings bank account 2, follewed by Credit Card

```{r}
mode_expense_count <- data %>%
  group_by(Mode) %>%
  summarise(Total = sum(Amount))

ggplot(mode_expense_count, aes(x = Mode, y = Total, fill = Mode)) +
  geom_col() +
  labs(title = "Number of Transactions per Mode Category",
       x = "Mode",
       y = "Number of Transactions") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```
#### Chart analysis
Chart shows there are two main Modes of transaction. Savings bank account 1, and Savings bank account 2, follewed by Credit Card


## Subcategory

**There are over 635 Subcategories, charting this variable wil not be helpful**


## Note

**Notes are sentences about the transaction, no charts**


# Bivariate Charts

## Amount and Date
```{r}
data %>% ggplot( aes(x=Date, y=Amount)) +
    geom_line() +
    # annotate(geom="text", x=as.Date("2017-01-01"), y=20089, 
    #          label="Bitcoin price reached 20k $\nat the end of 2017") +
    # annotate(geom="point", x=as.Date("2017-12-17"), y=20089, size=10, shape=21, fill="transparent") +
    geom_hline(yintercept=mean(data$Amount), color="orange", size=.5)
```

```{r}
# Calculate the mean of Amount for each Income.Expense group
mean_data <- data %>%
  group_by(Income.Expense) %>%
  summarise(mean_amount = mean(Amount))
ggplot(data, aes(x = Date, y = Amount, col = Income.Expense)) +
  geom_line() + 
  facet_wrap(~ Income.Expense, ncol = 1) +  # Facet by Income.Expense
  geom_hline(data = mean_data, aes(yintercept = mean_amount, color = Income.Expense), size = .5) +
  labs(title = "Time Series of Amount Over Date by Income/Expense",
       x = "Date",
       y = "Amount") 
```

## Amount and Mode
```{r}
mode_data_total <- data %>%
  group_by(Mode) %>%
  summarise(total_amount_sum = sum(Amount)) %>%
  arrange(total_amount_sum)

mode_data_count <- data %>%
  group_by(Mode) %>%
  summarise(total_amount_count = n()) %>%
  arrange(total_amount_count)

mode_data_total$Mode <- factor(mode_data_total$Mode, levels = mode_data_total$Mode)
mode_data_count$Mode <- factor(mode_data_count$Mode, levels = mode_data_total$Mode)

chart_1 <- mode_data_total %>% ggplot(aes(x = Mode, y = total_amount_sum, col = Mode, fill=Mode)) +
  geom_bar(stat="identity", alpha=.6, width=.4) +
  coord_flip() +
  labs(title = "Total Amount by Mode",
       x = "Mode",
       y = "Total Amount") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position = "none")

chart_2 <-mode_data_count %>% ggplot(aes(x = Mode, y = total_amount_count, col = Mode, fill=Mode)) +
  geom_bar(stat="identity", alpha=.6, width=.4) +
  coord_flip() +
  labs(title = "Count of Amount by Mode",
       x = "Mode",
       y = "Count Amount") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position = "none")

grid.arrange(chart_1, chart_2,ncol = 2)

```
#### Chart analysis
Chart shows Savings bank account has the highest largest sum of Amount Transactions, it also has the most transactions
Cash followed by credit card also have a high amount of transactions, 


## Amount and Category
```{r}
category_data_total <- data %>%
  group_by(Category) %>%
  summarise(total_amount_sum = sum(Amount)) %>%
  arrange(total_amount_sum)

category_data_count <- data %>%
  group_by(Category) %>%
  summarise(total_amount_count = n()) %>%
  arrange(total_amount_count)

category_data_total$Category <- factor(category_data_total$Category, levels = category_data_total$Category)
category_data_count$Category <- factor(category_data_count$Category, levels = category_data_total$Category)


chart_1 <- category_data_total %>% ggplot(aes(x = Category, y = total_amount_sum, col = Category, fill=Category)) +
  geom_bar(stat="identity", alpha=.6, width=.4) +
  coord_flip() +
  labs(title = "Total Amount by Category",
       x = "Category",
       y = "Total Amount") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position = "none")

chart_2 <-category_data_count %>% ggplot(aes(x = Category, y = total_amount_count, col = Category, fill=Category)) +
  geom_bar(stat="identity", alpha=.6, width=.4) +
  coord_flip() +
  labs(title = "Count of Amount by Category",
       x = "Category",
       y = "Count Amount") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position = "none")

grid.arrange(chart_1, chart_2,ncol = 2 )
```
## Amount and Subcategory



### Amount by Income.Expense Boxplot
This will explore the amount and Income.Expense variable in a box plot
``` {r}
plot(data$Amount ~ data$Income.Expense)
```
#### Chart analysis
There is a big difference between the amounts acorss the Income.Expense variable. 
Expense and Transfer-Out have on average small amounts, but also have several outlier values. 
Income has no outlier values but has a long tail, suggesting there is variety of amounts that 
are greater than the average, but on average the amount is small  

### Amount by Income only exploration
This will explore the Amount and Income category variable in a histogram
```{r}
income_amounts <- data %>% filter(Income.Expense == "Income")
summary(income_amounts$Amount)
```
Min is 2, and Max is 113376 
Q1 is 25, Q3 is 51562
Median is 2000 
Mean is 24339
there is considerable variance in the dat

### Amount by Income only Histogram
This will explore the Amount and Income category variable in a histogram
``` {r}
ggplot(income_amounts, aes(x = Amount)) +
  geom_histogram() + 
  labs(title = "Amount by Income Histogram",
       x = "Amount",
       y = "Count") 
```
#### Chart analysis
histogram is somewhat squewed showing two groups of income and one very lagre concentration at the start. 
The groups I can see is:
- small which are less than the median value of 2,000
- medium is between 2,000 and 10,000 
- large which are between 45,000 and 75,000

### Exploration of Amount groups
This will explore the two groups of income in histogram 

``` {r}
small_income_amounts <- income_amounts %>% filter(Amount <= 2000)

ggplot(small_income_amounts, aes(x = Amount)) +
  geom_histogram() + 
  labs(title = "Amount by Income Histogram",
       x = "Amount",
       y = "Count") 
```

``` {r}
medium_income_amounts <- income_amounts %>% filter((Amount > 2000) & (Amount < 10000))

ggplot(medium_income_amounts, aes(x = Amount)) +
  geom_histogram() + 
  labs(title = "Amount by Income Histogram",
       x = "Amount",
       y = "Count") 
```

``` {r}

large_income_amounts <- income_amounts %>% filter((Amount > 45000) & (Amount < 75000))

ggplot(large_income_amounts, aes(x = Amount)) +
  geom_histogram() + 
  labs(title = "Amount by Income Histogram",
       x = "Amount",
       y = "Count") 
```
#### Chart analysis
These dont provide any insights


# Multi-variate Charts

### Timeseries of Amount groups seperated by Income.Expense
This will chart amount and Income.Expense across the Date variable
```{r} 
# Create the time series plot with facets for each Income.Expense category
ggplot(data, aes(x = Date, y = Amount, col = Income.Expense)) +
  geom_line() + 
  facet_wrap(~ Income.Expense, ncol = 1) +  # Facet by Income.Expense
  labs(title = "Time Series of Amount Over Date by Income/Expense",
       x = "Date",
       y = "Amount") 
```

#### Chart analysis
The chart shows the amount of Income, Expense and Transfer-Out over time. Starting at 2015 up unitl the end of 2018
- The Expense group (red) has steady small amounts show regular pattern
- The Income group (green) goes up and down frequently. It rarly goes above 5000, but also seems to drop very low
- The Transfer-Out group seems to have a shorter range of amounts, starting in 2017. This sub type has some very 
 small amounts, and included some huge spikes in amounts

### Timeseries of Amount comparing Income and Expense
This will chart Income and Expense subtypes acorss the Date variable
```{r} 
income_and_expense_amounts <- data %>% filter(Income.Expense != "Transfer-Out")

# Create the time series plot with facets for each Income.Expense category
ggplot(income_and_expense_amounts, aes(x = Date, y = Amount, col = Income.Expense)) +
  geom_point() +
  geom_line() + 
  labs(title = "Time Series of Amount Over Date by Income/Expense",
       x = "Date",
       y = "Amount") 
```

#### Chart analysis
The chart shows the amount of Income and Expense and Transfer-Out over time. 
- Income is larger amount and less regular
- Expense is generally smaller amount and more regular

### Monthly Sum of Amount comparing Income.Expense

```{r} 

# Count the number of income transactions per month
income_freq <- data %>%
  group_by(MonthYear, Year, Income.Expense) %>%
  summarise(TotalAmount = sum(Amount))

head(income_freq)
# Plot the frequency of income transactions per month
ggplot(income_freq, aes(x = MonthYear, y = TotalAmount,col = Year, group = Income.Expense)) +
  geom_point() + 
   geom_line() + 
  facet_wrap(~ Income.Expense, ncol = 1) +
  labs(title = "Frequency of Income Transactions per Month",
       x = "Month-Year",
       y = "Number of Income Transactions") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
#### Chart analysis
The month shows a general constant line in Income. 
Expense has a less uniform line

### Quarterly Sum of Amount comparing Income.Expense
```{r} 

# Count the number of income transactions per month
season_income_freq <- data %>%
  group_by(SeasonYear, Year, Income.Expense) %>%
  summarise(TotalAmount = sum(Amount))

head(season_income_freq)
# Plot the frequency of income transactions per month
ggplot(season_income_freq, aes(x = SeasonYear, y = TotalAmount,col = Year, group = Income.Expense)) +
  geom_point() + 
   geom_line() + 
  facet_wrap(~ Income.Expense, ncol = 1) +
  labs(title = "Frequency of Income Transactions per Month",
       x = "Month-Year",
       y = "Number of Income Transactions") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
#### Chart analysis
The month shows a general constant line in Income. 
Expense has a less uniform line

### Weekly Sum of Amount comparing Income.Expense acorss years
```{r} 


# Count the number of income transactions per week
week_income_freq <- data %>%
  group_by(WeekYear, Year, Season, Month, Income.Expense) %>%
  summarise(TotalAmount = sum(Amount))

head(week_income_freq)
# Plot the frequency of income transactions per week
ggplot(week_income_freq, aes(x = WeekYear, y = TotalAmount,col = Year, group = Income.Expense)) +
  geom_point() + 
   geom_line() + 
  facet_wrap(~ Income.Expense, ncol = 1) +
  labs(title = "Frequency of Income Transactions per Week",
       x = "week-Year",
       y = "Number of Income Transactions") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(breaks = week_income_freq$WeekYear[seq(1, nrow(week_income_freq), by = 12)])
```
#### Chart analysis
The chart shows the **weekly sum of amounts** of Income and Expense and Transfer-Out over time. 
- Expense is generally smaller amount and more regular
- Income is larger amount and less regular
- Transfer-Out is over a shorter period and is composed of small and large amounts
- Both Expense and Income is stationary

### Same chart acorss seasons
```{r} 
# Plot the frequency of income transactions per week
ggplot(week_income_freq, aes(x = WeekYear, y = TotalAmount,col = Season, group = Income.Expense)) +
  geom_point() + 
   geom_line() + 
  facet_wrap(~ Income.Expense, ncol = 1) +
  labs(title = "Frequency of Income Transactions per Week",
       x = "week-Year",
       y = "Number of Income Transactions") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(breaks = week_income_freq$WeekYear[seq(1, nrow(week_income_freq), by = 12)])
```
#### Chart analysis
These dont provide any insights

### Same chart acorss months
```{r} 
# Plot the frequency of income transactions per month
ggplot(week_income_freq, aes(x = WeekYear, y = TotalAmount,col = Month, group = Income.Expense)) +
  geom_point() + 
   geom_line() + 
  facet_wrap(~ Income.Expense, ncol = 1) +
  labs(title = "Frequency of Income Transactions per Week",
       x = "week-Year",
       y = "Number of Income Transactions") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(breaks = week_income_freq$WeekYear[seq(1, nrow(week_income_freq), by = 12)])
```
#### Chart analysis
In both the Expense and income charts there appears to be a spike in amount in the months around the end of the year and January

### Further Exploration of Income.Expense data
First I will take a better look at the Transfer out data. 
```{r}
transfer_out_data <- data %>%
  filter(Income.Expense == "Transfer-Out")

str(transfer_out_data)
```
There are 160 Transfer-Out values in the data set. 

```{r}
summary(transfer_out_data)
```
The amounts range from 42.88 to 250,000, with an average of 11067

### Comparison of Income.Expense distributions 
Next I would like to get a better idea of the proportion the Transfer-Out value has in the dataset
```{r}
# Calculate the total amount per Income.Expense category
total_amount_per_category <- data %>%
  group_by(Income.Expense) %>%
  summarise(Total_Amount = sum(Amount))

#Create the bar chart for total amount per Income.Expense category
incomeExpenseAmountChart <-ggplot(total_amount_per_category, aes(x = Income.Expense, y = Total_Amount, fill = Income.Expense)) +
  geom_col(stat = "identity") +
  geom_text(aes(label = round(Total_Amount, 0)), 
            position = position_stack(vjust = 0.5), 
            color = "black", 
            size = 6) +
  labs(title = "Chart 1", subtitle = "Total Amount per Income/Expense Category",
       x = "Income/Expense",
       y = "Total Amount") +
  coord_flip()

  # Create a new column Is_Income
fundsIn_amount_chart <- data %>%
  mutate(FundsIn = Income.Expense == "Income") %>%
  group_by(Income.Expense, FundsIn) %>%
  summarise(Total_Amount = sum(Amount), )

  # Create the bar chart for Is_Income
incomeExpenseIsIncomeChart <- ggplot(fundsIn_amount_chart, aes(x = FundsIn, y = Total_Amount, fill = Income.Expense)) +
  geom_col(stat = "identity") +
    geom_text(aes(label = round(Total_Amount, 0)), 
            position = position_stack(vjust = 0.5), 
            color = "black", 
            size = 6) +
  labs(title = "Chart 2", subtitle = "Total Amount of funds in and out per Income/Expense Category",
       x = "Funds in and out",
       y = "Count")+
  coord_flip() 

percentage_per_category <- data %>%
  group_by(Income.Expense) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = (Count / sum(Count)) * 100)

# # Create the pie chart for percentage of each Income.Expense type
incomeExpensePercentageChart <- ggplot(percentage_per_category, aes(x = Income.Expense, y = Percentage, fill = Income.Expense)) +
  geom_col(stat = "identity") +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), 
            position = position_stack(vjust = 0.5), 
            color = "black", 
            size = 5) +
  labs(title = "Chart 3", subtitle = "Percentage of Income/Expense Types in Dataset",
       x = "Income/Expense",
       y = "Percentage (%)")+
  coord_flip() 



# # Combine the plots
grid.arrange(incomeExpenseAmountChart, incomeExpenseIsIncomeChart, incomeExpensePercentageChart, nrow = 3)

```

#### Chart analysis 
The above charts show three different horizontal Charts
 - Chart 1 shows that the sum of all income is the largest when compared against Expense and transfer-out. 
 - Chart 2 shows that the combined sums of Expense and transfer-out is larger than income, suggesting the 
 person is spending more than they are making. 
 - Chart 3 shows the total percentage of rows in the table per Income.Expense variable

## Categories and Income.Expense

### Categories across Income.Expense
Next I will explore the Categories across Income.Expense Varibale to get a better understanding of the Income.Expenses
```{r}
# Summarize the total amount per Category
category_amount_summary <- data %>%
  group_by(Category, Income.Expense) %>%
  summarise(Total_Amount = sum(Amount), .groups = 'drop') %>%
  select(Total_Amount, Income.Expense, Category) %>%
  arrange(desc(Total_Amount))

expenseCategoryAmount <- ggplot(category_amount_summary %>%
  filter(Income.Expense == "Expense"), aes(x = Category, y = Total_Amount, fill = Category)) +
  geom_col() +
  labs(title = "Expense", subtitle = "Expenses Total Amount per Category",
       x = "Category",
       y = "Total Amount") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels to 45 degrees
    plot.title = element_text(hjust = 0.5)
  )

incomeCategoryAmount <- ggplot(category_amount_summary %>%
  filter(Income.Expense == "Income"), aes(x = Category, y = Total_Amount, fill = Category)) +
  geom_col() +
  labs(title = "Income", subtitle = "Income Total Amount per Category",
       x = "Category",
       y = "Total Amount") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels to 45 degrees
    plot.title = element_text(hjust = 0.5)
  )

transferOutCategoryAmount <- ggplot(category_amount_summary %>%
  filter(Income.Expense == "Transfer-Out"), aes(x = Category, y = Total_Amount, fill = Category)) +
  geom_col() +
  labs(title = "Transfer-Out", subtitle = "Transfer-Out Total Amount per Category",
       x = "Category",
       y = "Total Amount") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels to 45 degrees
    plot.title = element_text(hjust = 0.5)
  )

grid.arrange(expenseCategoryAmount, incomeCategoryAmount, transferOutCategoryAmount, nrow = 3)
```

#### Chart analysis 
The large spike in the Income salary category apart that is Income.Expense is the largest amount

### Check if categories overlap
```{r}

data %>%
  group_by(Category) %>%
  summarise(Unique_Income_Expense = n_distinct(Income.Expense)) %>%
  filter(Unique_Income_Expense > 1)
```

There is some overlap of categories across income.expense types:
-   Other, in 2 categories
-   Public Provident Fund, in 2 categories
-   Recurring Deposit, in 2 categories

### Timeseries of Amount in Income comparing acorss categories
```{r}

# Count the number of income transactions per month
income_catergoy_freq <- data %>%
  filter(Income.Expense == "Income") %>%
  group_by(Category, MonthYear, Year) %>%
  summarise(TotalAmount = sum(Amount))

# Plot the frequency of income transactions per month
ggplot(income_catergoy_freq, aes(x = MonthYear, y = TotalAmount,col = Year, group = Category)) +
  geom_point() + 
   geom_line() + 
  facet_wrap(~ Category, ncol = 1) +
  labs(title = "Frequency of Income Transactions acorss Categories per Month",
       x = "Month-Year",
       y = "Number of Income Transactions") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Chart analysis 
Salary is the only Catergoy of Income that spans most of the duration of the dataset.
This also has seasonal dimension with regalar monthly amount, a slight trend upwards and very little variance
**will look at sub categories of salary category in further detail**

### Timeseries of Amount of salary acorss years
```{r}

income_salary_freq <- data %>%
  filter(Income.Expense == "Income" & Category == "Salary") %>%
  group_by(Month, Year) %>%
  summarise(TotalAmount = sum(Amount))

print(income_salary_freq)

ggplot(income_salary_freq, aes(x = Month, y = TotalAmount, color = factor(Year), group = factor(Year))) +
  geom_point() + 
  geom_line() + 
  facet_wrap(~ Year, ncol = 1) +
  labs(title = "Yearly Salary Transactions Per Month",
       x = "Month",
       y = "Salary Transactions Amount",
       color = "Year") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


#### Chart analysis 

### Timeseries of Amount of expense acorss years
```{r}

expense_freq <- data %>%
  filter(Income.Expense == "Expense") %>%
  group_by(Month, Year) %>%
  summarise(TotalAmount = sum(Amount))


ggplot(expense_freq, aes(x = Month, y = TotalAmount, color = factor(Year), group = factor(Year))) +
  geom_point() + 
  geom_line() + 
  facet_wrap(~ Year, ncol = 1) +
  labs(title = "Yearly Expense Transactions Per Month",
       x = "Month",
       y = "Number of Expense Transactions",
       color = "Year") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Timeseries of Amount Salary comparing acorss subcategories - No Chart
```{r}
salary_subcategory_freq <- data %>%
  filter(Category == "Salary") %>%
  group_by(Subcategory, MonthYear, Year) %>%
  summarise(TotalAmount = sum(Amount))

print(any(salary_subcategory_freq$Subcategory != ""))
```

#### analysis 
Salary has no sub categories

### Timeseries of Amount in Expense comparing acorss categories
```{r}

# Count the number of expense transactions per week
expense_catergoy_freq <- data %>%
  filter(Income.Expense == "Expense") %>%
  
  group_by(Category, WeekYear, Year) %>%
  summarise(TotalAmount = sum(Amount))

head(income_freq)
# Plot the frequency of expense transactions per week
ggplot(expense_catergoy_freq, aes(x = WeekYear, y = TotalAmount,col = Year, group = Category)) +
  geom_point() + 
   geom_line() + 
  facet_wrap(~ Category, ncol = 1) +
  labs(title = "Frequency of Expense Transactions acorss Categories per Week",
       x = "Week-Year",
       y = "Number of Income Transactions") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Chart analysis 
Unlike Income, the Expense type has there are many Categories that have transactions throughout the year. 

### Analysis of Expense Categories and subcategories
```{r}

# Count the number of expense transactions per week
expense_catergoy_freq <- data %>%
  filter((Income.Expense == "Expense") & (Subcategory != "")) %>%
  group_by(Category, Subcategory) %>%
  summarise(TotalAmount = sum(Amount))

head(expense_catergoy_freq)

ggplot(expense_catergoy_freq, aes(x = Category, y = TotalAmount, fill = Subcategory)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Expense Amount by Category and Subcategory",
       x = "Category",
       y = "Total Amount",
       fill = "Subcategory") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Expense Prediction Model

## clean Expense data
```{r}
data %>%
#  filter(Income.Expense == "Expense") %>%
 ggplot( aes(x = Income.Expense, y = Amount)) +
  geom_boxplot() +
  labs(title = "Boxplot of Values by Group",
       x = "Group",
       y = "Value") +
  theme_minimal()

```

```{r}
expense_data <- data %>%
  filter(Income.Expense == "Expense")

Q1 <- quantile(expense_data$Amount, 0.25)
Q3 <- quantile(expense_data$Amount, 0.75)
IQR_value <- Q3 - Q1

lower_bound <- Q1 - 1.5 * IQR_value
upper_bound <- Q3 + 1.5 * IQR_value

cleaned_data <- expense_data %>%
  filter(Amount >= lower_bound & Amount <= upper_bound)

# Plot the cleaned data
ggplot(cleaned_data, aes(x = Income.Expense, y = Amount)) +
  geom_boxplot() +
  labs(title = "Boxplot of Values by Group (Outliers Removed)",
       x = "Group",
       y = "Value") 
```

```{r}
hist(cleaned_data$Amount)
```

```{r}
cleaned_data$Amount_log <- log(cleaned_data$Amount)
hist(cleaned_data$Amount_log)
```

```{r}
# Plot the cleaned data
ggplot(cleaned_data, aes(x = Income.Expense, y = Amount_log)) +
  geom_boxplot() +
  labs(title = "Boxplot of Values by Group (Outliers Removed)",
       x = "Group",
       y = "Value") +
  theme_minimal()
```

## TSLM weekly expense model

### Expense Data Preperation 
Createing a weekly expense dataframe
```{r}
weekly_expense <- cleaned_data %>%
  group_by(weeknumber) %>%
  summarise(
    Total_Expense = sum(Amount_log),
    year_number,
    weeknumber,
    Season,
    Month,
    Year, 
    Week, 
    Date
    )
```

### Weekly expense time seires
```{r}
  # Determine the start year and month
start_year <- min(weekly_expense$year_number)
start_week <- min(weekly_expense$weeknumber)

    # Convert to time series object
weekly_expense_ts <- ts(weekly_expense$Total_Expense, start = c(start_year, start_week), frequency = 52)
```

### Create Time series model
```{r}

# Fit tslm model with additional features
weekly_model <- tslm(weekly_expense_ts ~ trend +  year_number+
    weeknumber+
    Season+
    Month+
    Year+
    Week , data=weekly_expense)
summary(weekly_model)
```


### Plot Actual vs Predicted
```{r}
# Create a data frame for actual vs predicted values
comparison_df <- data.frame(
  Week = 1:length(weekly_expense$Total_Expense),
  Actual = weekly_expense$Total_Expense,
  Predicted = fitted(weekly_model),
  Year = weekly_expense$Year
)

# Plot actual vs predicted values
ggplot(comparison_df, aes(x = Week)) +
  geom_line(aes(y = Actual, color = "Actual")) +
  geom_line(aes(y = Predicted, color = "Predicted")) +
  facet_wrap(~ Year, ncol = 1) +
  labs(title = "Actual vs Predicted Monthly Expenses",
       x = "Week",
       y = "total Amount") 
```

## simplified TSLM weekly expense model
```{r}
# Fit a simplified tslm model
simplified_model <- tslm(weekly_expense_ts ~ Season + Month + Week, data = weekly_expense)

# Summary of the simplified model
summary(simplified_model)

```

### Plot Simplified Actual vs Simplified Predicted
```{r}
# Create a data frame for actual vs predicted values
comparison_df <- data.frame(
  Week = 1:length(weekly_expense$Total_Expense),
  Actual = weekly_expense$Total_Expense,
  Predicted = fitted(simplified_model),
  Year = weekly_expense$Year
)

# Plot actual vs predicted values
ggplot(comparison_df, aes(x = Week)) +
  geom_line(aes(y = Actual, color = "Actual")) +
  geom_line(aes(y = Predicted, color = "Predicted")) +
  facet_wrap(~ Year, ncol = 1) +
  labs(title = "Simplified Model Actual vs Predicted Monthly Expenses",
       x = "Week",
       y = "total Amount") 
```


## ARIMA Expense model 
### attempt 1
```{r}

arima_model <- auto.arima(weekly_expense_ts)

summary(arima_model)

```

```{r}
forecasted_values <- forecast(arima_model, h = 52)  
plot(forecasted_values)
```

### attempt 2
```{r}

arima_model_2 <- auto.arima(weekly_expense_ts, seasonal = FALSE)

summary(arima_model_2)

```

```{r}
forecasted_values_2 <- forecast(arima_model_2, h = 52)  
plot(forecasted_values_2)
```

### attempt 3

```{r}

arima_model_3 <- auto.arima(weekly_expense_ts, seasonal = FALSE, stepwise=FALSE, approximation=FALSE)

summary(arima_model_3)

```

```{r}
forecasted_values_3 <- forecast(arima_model_3, h = 52)  
plot(forecasted_values_3)
```



### attempt 4

```{r}
ggAcf(weekly_expense_ts)
```

```{r}
ggPacf(weekly_expense_ts)
```
```{r}

arima_model_4 <-Arima(weekly_expense_ts, order=c(0, 1, 3))

summary(arima_model_4)

```

```{r}
forecasted_values_4 <- forecast(arima_model_4, h = 52)  
plot(forecasted_values_4)
```
## Multi Linear Regression with One Hot encoding


### encoding data
```{r}
encoding_catergoris <-  cleaned_data %>%
  summarise(
    Mode,
    Category,
    Subcategory, 
    Year, 
    Month,
    MonthYear,
    Season, 
    SeasonYear,
    Week, 
    WeekYear
    )

#define one-hot encoding function
dummy <- dummyVars(" ~ .", data=encoding_catergoris)

#perform one-hot encoding on data frame
encoded_data <- data.frame(predict(dummy, newdata=encoding_catergoris))

```

```{r}
#view final data frame
str(encoded_data)
```

```{r}
# combine encoded with original dataframe
combined_data <- bind_cols(cleaned_data, encoded_data)


str(combined_data)
categorical_encoded_expense_data <- combined_data %>%
  select(
    -Income.Expense,
    -Date, 
    -Mode, 
    -Category,
    -Subcategory, 
    -Year, 
    -Note,
    -Currency,
   -Month,
    -MonthYear,
    -WeekYear, 

     -year_number, 
     -weeknumber   )
```

### model
```{r}

# Ensure the target variable is numeric

# Split the data
set.seed(123)  # For reproducibility
train_indices <- sample(seq_len(nrow(categorical_encoded_expense_data)), size = 0.8 * nrow(categorical_encoded_expense_data))
train_data <- categorical_encoded_expense_data[train_indices, ]
test_data <- categorical_encoded_expense_data[-train_indices, ]

source_data <- cleaned_data[-train_indices, ]

# Build the model
model <- lm(Amount_log ~ ., data = train_data)

summary(model)
```
Many if the Coefficients have a high p-value, so later on I will remove these


```{r}
# Make predictions
predictions <- predict(model, newdata = test_data)
```

```{r}

# Create a data frame for actual vs predicted values
comparison_df <- data.frame(
  Actual = test_data$Amount_log,
  Predicted = predictions,
  Week=source_data$Week,
  Year=source_data$Year
)

# Coefficients of the model
coef(model)
```

```{r}

ggplot(comparison_df, aes(x = Actual, y = Predicted)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Actual vs Predicted Expenses",
       x = "Actual Expenses",
       y = "Predicted Expenses")

```

### 
```{r}
ggplot(comparison_df, aes(x = Week)) +
  geom_line(aes(y = Actual, color = "Actual")) +
  geom_line(aes(y = Predicted, color = "Predicted")) +
  facet_wrap(~ Year, ncol = 1) +
  labs(title = "Actual vs Predicted Expenses",
       x = "Actual Expenses",
       y = "Predicted Expenses")
```

<!-- plot residuals  -->
```{r}
plot(model, which = 1)
```

```{r}
plot(model, which = 2)
```

### simplified linear model
```{r}
# Step 1: Extract the p-values
p_values <- summary(model)$coefficients[, "Pr(>|t|)"]

# Step 2: Identify insignificant features
# Let's use a threshold of 0.05
insignificant_features <- names(p_values[p_values <= 0.05])

# Step 3: Remove insignificant features and create a new model
# First, create a formula excluding the insignificant features
formula <- as.formula(paste("Amount_log ~ ", paste(insignificant_features, collapse = " + ")))

# Build the new model
new_model <- lm(formula, data = train_data)

# Summary of the new model
summary(new_model)
```
New model has better coefficients

```{r}
# Make predictions
new_predictions <- predict(new_model, newdata = test_data)
```

```{r}
# Make predictions
# predictions <- predict(model, newdata = test_data)

# Calculate performance metrics
# mse <- mean((test_data$Amount - predictions)^2)
# rmse <- sqrt(mse)
# mae <- mean(abs(test_data$Amount - predictions))
# r_squared <- summary(model)$r.squared

# list(MSE = mse, RMSE = rmse, MAE = mae, R_squared = r_squared)

# Create a data frame for actual vs predicted values
comparison_df <- data.frame(
  Actual = test_data$Amount_log,
  Predicted = new_predictions,
  Week=source_data$Week,
  Year=source_data$Year
)

# Coefficients of the model
coef(new_model)
```

```{r}

comparison_df$Color <- ifelse(comparison_df$Actual > comparison_df$Predicted, "blue", "green")

ggplot(comparison_df, aes(x = Actual, y = Predicted)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Actual vs Predicted Expenses",
       x = "Actual Expenses",
       y = "Predicted Expenses")

```

### 
```{r}
ggplot(comparison_df, aes(x = Week)) +
  geom_line(aes(y = Actual, color = "Actual")) +
  geom_line(aes(y = Predicted, color = "Predicted")) +
  facet_wrap(~ Year, ncol = 1) +
  labs(title = "Actual vs Predicted Expenses",
       x = "Actual Expenses",
       y = "Predicted Expenses")
```

<!-- plot residuals  -->
```{r}
plot(new_model, which = 1)
```

```{r}
plot(new_model, which = 2)
```


# Income Salary Prediction Model

## clean Income data
```{r}
income_salary_data <- data %>%
  filter(Income.Expense == "Income" & Category == "Salary")

Q1 <- quantile(income_salary_data$Amount, 0.25)
Q3 <- quantile(income_salary_data$Amount, 0.75)
IQR_value <- Q3 - Q1

lower_bound <- Q1 - 1.5 * IQR_value
upper_bound <- Q3 + 1.5 * IQR_value

cleaned_salary_data <- income_salary_data %>%
  filter(Amount >= lower_bound & Amount <= upper_bound)

# Plot the cleaned data
ggplot(cleaned_salary_data, aes(x = Income.Expense, y = Amount)) +
  geom_boxplot() +
  labs(title = "Boxplot of Salary Income Amount",
       x = "Group",
       y = "Value") +
  theme_minimal()
```

```{r}
hist(cleaned_salary_data$Amount)
```

```{r}
cleaned_salary_data$Amount_log <- log(cleaned_salary_data$Amount)
hist(cleaned_salary_data$Amount_log)
```

```{r}
# Plot the cleaned data
ggplot(cleaned_salary_data, aes(x = Income.Expense, y = Amount_log)) +
  geom_boxplot() +
  labs(title = "Boxplot of Values by Group (Outliers Removed)",
       x = "Group",
       y = "Value") +
  theme_minimal()
```

## TSLM Monthly Income Salary  model

### Income Salary  Data Preperation 
Createing a weekly expense dataframe
```{r}
monthly_salay_income <- cleaned_salary_data %>%
  group_by(MonthYear) %>%
  summarise(
    Total_Income = sum(Amount_log),
    year_number,
    weeknumber,
    Season,
    Month,
    Year, 
    Week, 
    Date
    )
```

### Monthly Income Salary  time seires
```{r}
  # Determine the start year and month
start_year <- min(monthly_salay_income$Year)
start_week <- min(monthly_salay_income$Month)

monthly_salay_income_ts <- ts(monthly_salay_income$Total_Income, start = c(start_year, start_week), frequency = 12)
```

### Create Time series model
```{r}

# Fit tslm model with additional features
monthly_salay_income_model <- tslm(monthly_salay_income_ts ~ trend +  year_number+
    weeknumber+
    Season+
    Month+
    Year+
    Week , data=monthly_salay_income)
summary(monthly_salay_income_model)
```


### Plot Actual vs Predicted
```{r}
# Create a data frame for actual vs predicted values
comparison_df <- data.frame(
  Month = 1:length(monthly_salay_income$Total_Income),
  Actual = monthly_salay_income$Total_Income,
  Predicted = fitted(monthly_salay_income_model),
  Year = monthly_salay_income$Year
)

# Plot actual vs predicted values
ggplot(comparison_df, aes(x = Month)) +
  geom_line(aes(y = Actual, color = "Actual")) +
  geom_line(aes(y = Predicted, color = "Predicted")) +
  facet_wrap(~ Year, ncol = 1) +
  labs(title = "Actual vs Predicted Monthly Expenses",
       x = "Week",
       y = "total Amount") 
```

## simplified TSLM monthly Income Salary  model
```{r}
# Fit a simplified tslm model
simplified_model <- tslm(monthly_salay_income_ts ~ trend + year_number+
    Season+
    Month+
    Week , data=monthly_salay_income)

# Summary of the simplified model
summary(simplified_model)

```

### Plot Simplified Actual vs Simplified Predicted
```{r}
# Create a data frame for actual vs predicted values
comparison_df <- data.frame(
  Month = 1:length(monthly_salay_income$Total_Income),
  Actual = monthly_salay_income$Total_Income,
  Predicted = fitted(simplified_model),
  Year = monthly_salay_income$Year
)

# Plot actual vs predicted values
ggplot(comparison_df, aes(x = Month)) +
  geom_line(aes(y = Actual, color = "Actual")) +
  geom_line(aes(y = Predicted, color = "Predicted")) +
  facet_wrap(~ Year, ncol = 1) +
  labs(title = "Simplified Model Actual vs Predicted Monthly Expenses",
       x = "Week",
       y = "total Amount") 
```


## ARIMA Income Salary  model 
### attempt 1
```{r}

income_salary_arima_model <- auto.arima(monthly_salay_income_ts)

summary(income_salary_arima_model)

```

```{r}
forecasted_income_salary <- forecast(income_salary_arima_model, h = 52)  
plot(forecasted_income_salary)
```

### attempt 2
```{r}

arima_model_2 <- auto.arima(weekly_expense_ts, seasonal = FALSE)

summary(arima_model_2)

```

```{r}
forecasted_values_2 <- forecast(arima_model, h = 52)  
plot(forecasted_values_2)
```