---
title: "Data Cleaning"
author: "Padraic O'Hora"
date: "19/11/2024"
output: pdf_document
---

``` {r include=FALSE}
data <- read.csv("./Daily_Household_Transactions.csv")

head(data)



```


``` {r}
str(data)
```

It appears some data types are not correct. I will now change the types
``` {r}
length(unique(data$Mode)) # 12
length(unique(data$Category)) #50
length(unique(data$Subcategory)) #91
length(unique(data$Income.Expense)) #3 
length(unique(data$Currency)) #1

data$Mode <- as.factor(data$Mode)
data$Category <- as.factor(data$Category)
data$Subcategory <- as.factor(data$Subcategory)
data$Income.Expense <- as.factor(data$Income.Expense)

data$Date <- parse_date_time(data$Date, orders = c("d/m/Y HMS", "d/m/Y"))


```

``` {r}
summary(is.na(data))
```

No na values in the data set
Next I will perform univariate analysis

``` {r}
hist(data$Amount)
```

``` {r}
boxplot(data$Amount)
```

both charts show that the amount variable has a considerable amount of outlyer values

``` {r}
data$Amount_log <- log(data$Amount)
hist(data$Amount_log)
```

``` {r}
boxplot(data$Amount_log)
```

This looks better, but I would lik eto find what other variable influence the amount variable
``` {r}
plot(data$Amount ~ data$Income.Expense)
```
There is a big difference between the amounts depending on the Income.Expense Category

``` {r}
# Load the ggplot2 package
library(ggplot2)

# Create the time series plot
ggplot(data, aes(x = Date, y = Amount)) +
  geom_line() + 
  labs(title = "Time Series of Amount Over Date",
       x = "Date",
       y = "Amount") 

```
The plot seems to show some reoccurence in pattern. 

Lets create the same plot but seperate it by income.Expense
```{r} 
# Create the time series plot with facets for each Income.Expense category
ggplot(data, aes(x = Date, y = Amount, col = Income.Expense)) +
  geom_line() + 
  facet_wrap(~ Income.Expense, ncol = 1) +  # Facet by Income.Expense
  labs(title = "Time Series of Amount Over Date by Income/Expense",
       x = "Date",
       y = "Amount") 
```

The Expense and income variables show transcations throughout the timeframe. 
Whereas the Transfer out takes place in small timeframe. This leads me to believe 
the transfer out data is not relevean here and could be skewing some of the graphs. 

First I will take a better look at the Transfer out data. 
```{r}
library(dplyr)

transfer_out_data <- data %>%
  filter(Income.Expense == "Transfer-Out")

str(transfer_out_data)
```
There are 160 Transfer-Out values in the data set. 

```{r}
summary(transfer_out_data)
```
The amounts range from 42.88 to 250,000, with an average of 11067

Next I would like to get a better idea of the proportion the Transfer-Out value has in the dataset
```{r}
library(gridExtra)

# Calculate the total amount per Income.Expense category
total_amount_per_category <- data %>%
  group_by(Income.Expense) %>%
  summarise(Total_Amount = sum(Amount))

#Create the bar chart for total amount per Income.Expense category
incomeExpenseAmountChart <-ggplot(total_amount_per_category, aes(x = Income.Expense, y = Total_Amount, fill = Income.Expense)) +
  geom_col(stat = "identity") +
  geom_text(aes(label = round(Total_Amount, 0)), 
            position = position_stack(vjust = 0.5), 
            color = "black", 
            size = 6) +
  labs(title = "Chart 1", subtitle = "Total Amount per Income/Expense Category",
       x = "Income/Expense",
       y = "Total Amount") +
  coord_flip()

  # Create a new column Is_Income
fundsIn_amount_chart <- data %>%
  mutate(FundsIn = Income.Expense == "Income") %>%
  group_by(Income.Expense, FundsIn) %>%
  summarise(Total_Amount = sum(Amount), )

  # Create the bar chart for Is_Income
incomeExpenseIsIncomeChart <- ggplot(fundsIn_amount_chart, aes(x = FundsIn, y = Total_Amount, fill = Income.Expense)) +
  geom_col(stat = "identity") +
    geom_text(aes(label = round(Total_Amount, 0)), 
            position = position_stack(vjust = 0.5), 
            color = "black", 
            size = 6) +
  labs(title = "Chart 2", subtitle = "Total Amount of funds in and out per Income/Expense Category",
       x = "Funds in and out",
       y = "Count")+
  coord_flip() 

percentage_per_category <- data %>%
  group_by(Income.Expense) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = (Count / sum(Count)) * 100)

# # Create the pie chart for percentage of each Income.Expense type
incomeExpensePercentageChart <- ggplot(percentage_per_category, aes(x = Income.Expense, y = Percentage, fill = Income.Expense)) +
  geom_col(stat = "identity") +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), 
            position = position_stack(vjust = 0.5), 
            color = "black", 
            size = 5) +
  labs(title = "Chart 3", subtitle = "Percentage of Income/Expense Types in Dataset",
       x = "Income/Expense",
       y = "Percentage (%)")+
  coord_flip() 



# # Combine the plots
grid.arrange(incomeExpenseAmountChart, incomeExpenseIsIncomeChart, incomeExpensePercentageChart, nrow = 3)

```
The above chart shows two bar charts



```{r}
percentage_per_category <- data %>%
  group_by(Income.Expense) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = (Count / sum(Count)) * 100)

# Create the bar chart for percentage of each Income.Expense type
ggplot(percentage_per_category, aes(x = Income.Expense, y = Percentage, fill = Income.Expense)) +
  geom_bar(stat = "identity") +
  labs(title = "Percentage of Income/Expense Types in Dataset",
       x = "Income/Expense",
       y = "Percentage (%)") 
```